Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-04-24 WSSSPE1_report/WSSSPE1_summary_report.tex
Some minor corrections of typos, sentences
Added in a few additional tool examples where they seemed appropriate.

Hanwell et al.~\cite{Hanwell_WSSSPE} reflect on the 15-year history of
open source software development at \toolname{Kitware}.  In
particular, they focus on their success at growing their community of
users through enabling multiple channels of communication, directly
reaching out to individuals, and lowering the barrier to entry for
contributions. This involves providing clear, test-oriented and
review-based mechanisms for evaluating contributions, permissive
licenses, and a service-based model for sustaining development.  This
model enables \toolname{Kitware} to review both public funding, as
well as private funding to support improvements and targeted
developments of the software.

@@

What is, or should be meant by ``sustainability'' in the context of
software came up in many different parts of the workshop, specifically
in the first keynote (\S\ref{sec:keynote1}), the Developing and
Maintaining Software panel (\S\ref{sec:devel}), and the Policy panel
(\S\ref{sec:policy}). It quickly became clear that at present there is
no consensus among the community, whether within or accross
disciplines, on what this definition should be, and that a variety of
different defintiions were being assumed, used, or sometime expressly
articulated by contributors and attendees. However, some concepts,
particularly relating sustainability to change over time, were also
evidently held in common. This common notion is, for example, captured
in the definition used by the UK's Scientific Software Sustainability
Institute, quoted in~\cite{Venters_WSSSPE}: ``software you use today
will be available---and continue to be improved and supported---in the
future''. Pierce et al~\cite{Pierce_WSSSPE} express this idea as
software that continues to serve its users.

@@

Attendees suggested that having a definition of sustainability on
which the community can agree is key.  A related question that was
raised is what the goal of sustainability should be, with a wide range
of possible answers, including more reproducible science, software
persistence, and quality.  And given a goal of sustainability, how
would success in achieving it be measured?  How would one know that a
piece of software has reached sustainability? Participants pointed out
that for truly sustainable software there should be no endpoint at
which sustainability can be claimed, because the software products
would continue to be used and useful beyond the single institution,
grant, and developer or development team that created them. This may
mean that sustainability needs to be addressed throughout the full
software life cycle.  It was also pointed out that software
sustainability is not isolated from other attributes of scientific
software and its use, such as usability, and provenance. Similarly,
the question was considered, albeit only briefly, whether proprietary
versus open-source license plays a role in the context softwara
sustainability. For example, should a project ensure that it uses an
OSI-approved license so that software products can outlive any single
entity if they remain important.

@@

The career path issue emerged repeatedly across panels because for
scientific software to be(come) sustainable, projects trying to create
sustainable software need to be able to recruit and retain software
developers trained in the various requisite software engineering
facets.  However, a career path in research means faculty at most
universitities, and as was noted repeatedly in discussions, faculty
are hired based on their scientific qualifications, not on their
software development skills or track record. Consequently, developing
special software development skills is unlikely to further a career in
science at a university, although national laboratories were
acknowledged as a different case. Loffler et
al.~\cite{Loffler_WSSSPE}, one of the papers in the Communities panel
(\S\ref{sec:community}), brought the problem to the point:
\begin{quote}
  ``The most severe problem for developers in most computational
  sciences currently is that while most of the work is done creating
  hopefully well-written, sustainable software, the academic success
  is often exclusively tied to the solution of the scientific problem
  the software was designed for. Tasks that from a software
  engineering standpoint are essential, e.g., high usability,
  well-written and updated documentation, or porting infrastructure to
  new platforms, are not rewarded within this system.'' ~\cite{Loffler_WSSSPE}
\end{quote}

@@

% Why are we classifying software like above?
This classification is motivated by our argument that the two kinds of
software projects wildly vary in factors such as scope, purpose and
usage. The development and management of each kind is significantly
different.  Consequently, the sustainability challenges faced by them
differ and must be treated separately. For instance, the challenges
faced by a gateway software development project such as
\toolname{CIPRES} or a visualization software product such as
\toolname{VisIT} are distinct to a niche software for \textit{ab
  initio} modeling and simulation such as \toolname{VASP} or
\toolname{Quantum Espresso}.

\subsection{Utility Software}
Software developed with a potentially wider audience and general
purpose usage in mind is utility software. Utility software typically
does not address fundamental research problems for a given scientific
domain. Examples are collaborative development frameworks such as
\toolname{GitHub} and \toolname{Bitbucket}, distributed workflow and
generic computing frameworks such as \toolname{Galaxy},
\toolname{nanoHUB}, \toolname{SimGrid}, \toolname{Swift},
\toolname{Globus} and \toolname{VisTrails}, and visualization
frameworks such as \toolname{VisIT}.

@@

One such example is the \toolname{Galaxy} project~\cite{Galaxy}. It
follows agile software development practices and implements standards
such as test-driven development and socialized bug managing practices
via \toolname{trello}. Galaxy \emph{histories} and \emph{toolshed}
offers easy community sharing of data and tools further promoting a
collaborative environment. The project very closely follows the
guidelines described in Carver and Thiruvathukal~\cite{Carver_WSSSPE}
and many from Prli\'{c} and Procter~\cite{Prlic_WSSSPE}. Many utility
software projects are often developed aiming better utilizing a
particular, new infrastructure and architecture, e.g.,
\toolname{MVAPICH}, \toolname{VisIT}. Similarly, to leverage the power
of emerging architectures such as accelerators, new code and libraries
are required. The experience of one such effort as described in
Ferenbaugh~\cite{Ferenbaugh_WSSSPE} which met with a limited success
but nonetheless with many invaluable lessons were learned about
influential cultural and technical aspects in sustainable software
development practices.

>>>>

Hanwell et al.~\cite{Hanwell_WSSSPE} reflect on the 15-year history of
open source software development at \toolname{Kitware}.  In
particular, they focus on their success at growing their community of
users through enabling multiple channels of communication, directly
reaching out to individuals, and lowering the barrier to entry for
contributions. This involves providing clear, test-oriented and
review-based mechanisms for evaluating contributions, permissive
licenses, and a service-based model for sustaining development.  This
model enables \toolname{Kitware} to receive both public funding, as
well as private funding to support improvements and targeted
developments of the software.

@@

What is, or should be meant by ``sustainability'' in the context of
software came up in many different parts of the workshop, specifically
in the first keynote (\S\ref{sec:keynote1}), the Developing and
Maintaining Software panel (\S\ref{sec:devel}), and the Policy panel
(\S\ref{sec:policy}). It quickly became clear that at present there is
no consensus among the community, whether within or across
disciplines, on what this definition should be, and that a variety of
different definitions were being assumed, used, or sometime expressly
articulated by contributors and attendees. However, some concepts,
particularly relating sustainability to change over time, were also
evidently held in common. This common notion is, for example, captured
in the definition used by the UK's Scientific Software Sustainability
Institute, quoted in~\cite{Venters_WSSSPE}: ``software you use today
will be available---and continue to be improved and supported---in the
future''. Pierce et al~\cite{Pierce_WSSSPE} express this idea as
software that continues to serve its users.

@@

Attendees suggested that having a definition of sustainability on
which the community can agree is key.  A related question that was
raised is what the goal of sustainability should be, with a wide range
of possible answers, including more reproducible science, software
persistence, and quality.  And given a goal of sustainability, how
would success in achieving it be measured?  How would one know that a
piece of software has reached sustainability? Participants pointed out
that for truly sustainable software there should be no endpoint at
which sustainability can be claimed, because the software products
would continue to be used and useful beyond the single institution,
grant, and developer or development team that created them. This may
mean that sustainability needs to be addressed throughout the full
software life cycle.  It was also pointed out that software
sustainability is not isolated from other attributes of scientific
software and its use, such as usability, and provenance. Similarly,
the question was considered, albeit only briefly, whether proprietary
versus open-source license plays a role in the context of software
sustainability. For example, should a project ensure that it uses an
OSI-approved license so that software products can outlive any single
entity if they remain important.

@@

The career path issue emerged repeatedly across panels because for
scientific software to be(come) sustainable, projects trying to create
sustainable software need to be able to recruit and retain software
developers trained in the various requisite software engineering
facets.  However, a career path in research means faculty at most
universities, and as was noted repeatedly in discussions, faculty
are hired based on their scientific qualifications, not on their
software development skills or track record. Consequently, developing
special software development skills is unlikely to further a career in
science at a university, although national laboratories were
acknowledged as a different case. Loffler et
al.~\cite{Loffler_WSSSPE}, one of the papers in the Communities panel
(\S\ref{sec:community}), brought the problem to the point:
\begin{quote}
  ``The most severe problem for developers in most computational
  sciences currently is that while most of the work is done creating
  hopefully well-written, sustainable software, the academic success
  is often exclusively tied to the solution of the scientific problem
  the software was designed for. Tasks that from a software
  engineering standpoint are essential, e.g., high usability,
  well-written and updated documentation, or porting infrastructure to
  new platforms, are not rewarded within this system.'' ~\cite{Loffler_WSSSPE}
\end{quote}

@@

% Why are we classifying software like above?
This classification is motivated by our argument that the two kinds of
software projects wildly vary in factors such as scope, purpose and
usage. The development and management of each kind is significantly
different.  Consequently, the sustainability challenges faced by them
differ and must be treated separately. For instance, the challenges
faced by a gateway software development project such as
\toolname{CIPRES} or visualization software products such as
\toolname{VisIT} or \toolname{ParaView} are distinct to a niche software for
\textit{ab initio} modeling and simulation such as \toolname{VASP} or
\toolname{Quantum Espresso}.

\subsection{Utility Software}
Software developed with a potentially wider audience and general
purpose usage in mind is utility software. Utility software typically
does not address fundamental research problems for a given scientific
domain. Examples are collaborative development frameworks such as
\toolname{GitHub} and \toolname{Bitbucket}, distributed workflow and
generic computing frameworks such as \toolname{Galaxy},
\toolname{nanoHUB}, \toolname{SimGrid}, \toolname{Swift},
\toolname{Globus} and \toolname{VisTrails}, and visualization
frameworks such as \toolname{VTK}, \toolname{VisIT}, and \toolname{ParaView}.

@@

One such example is the \toolname{Galaxy} project~\cite{Galaxy}. It
follows agile software development practices and implements standards
such as test-driven development and socialized bug managing practices
via \toolname{trello}. Galaxy \emph{histories} and \emph{toolshed}
offers easy community sharing of data and tools further promoting a
collaborative environment. The project very closely follows the
guidelines described in Carver and Thiruvathukal~\cite{Carver_WSSSPE}
and many from Prli\'{c} and Procter~\cite{Prlic_WSSSPE}. Many utility
software projects are often developed aiming better utilizing a
particular, new infrastructure and architecture, e.g.,
\toolname{MVAPICH}, \toolname{VisIT}, toolname{ParaView}. Similarly, to leverage the power
of emerging architectures such as accelerators, new code and libraries
are required. The experience of one such effort as described in
Ferenbaugh~\cite{Ferenbaugh_WSSSPE} which met with a limited success
but nonetheless with many invaluable lessons were learned about
influential cultural and technical aspects in sustainable software
development practices.
